The non-random assembly of indirect interactions in  \nplant-pollinator networks

### 2022-04-26 Manuscript/Main_text/Main_text_Journal_of_Animal_Ecol.Rmd
adding Nacho's edits

---
title: "The non-random assembly of functional motifs in  \nplant-pollinator networks"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[left]{lineno}
- \linenumbers
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
biblio-style: apalike
csl: functional-ecology.csl #funct. ecol. style from zotero styles
---

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of mutualistic networks. Detecting those relationships is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different plant and floral visitor groups on both trophic levels. Bees appear less frequently in specialised motif positions with high number of indirect interactions, while the rest of floral visitor groups are infrequent in generalised motif positions with low number of indirect interactions. All plant groups tend to be over-represented on specialised motif positions, except tall plant species with separated sexes (e.g., woody dioecious or monoecious species), which are more frequent on motif positions with low number of indirect interactions. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlights the non-random structure of the meso-scale on plant-pollinators networks and the association of different plant and floral visitor groups with certain motifs that involve different ecological roles at a macro-ecological scale.

Keywords: plant-pollinator networks, plant-pollinator interactions, network motifs, indirect interactions, plant traits, floral visitors

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions (i.e., the mediated effect between two species by a third species), such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs [small sub-networks observed within a given network which are often referred as its building blocks\; @milo2002] allows to consider both direct and indirect interactions [@simmons2019]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches at the node or at the full network level are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network motifs can capture both direct and indirect interactions, by analysing over- and under-representation of motifs we were able to explore in great detail the structure of interactions of plant-pollinator networks. We have explored for the first time the overall structure and representation of network motifs in real plant-pollinator networks at a macro-ecological scale and shown that the different motifs up to five nodes are consistently over- or under-represented in comparison with simulated networks. Despite the presence of multiple ecological processes (e.g., neutral effects, morphological matching or phenological overlap) shaping motif structure, we found a tendency for over-represented structures with more densely connected motifs (i.e., complete, fan and asymmetric complete). As suggested in @simmons2020, a higher proportion of these densely connected motifs could reflect that neutral effects are the predominant ecological process governing the structure of network motifs. Thus, as indicated for other network properties like nestedness at the macro-scale level [@krishna2008; @suweis2013], or interaction probability at the species level [@bartomeus2016], species relative abundances (i.e., neutral effects) may be also the main driver governing the meso-scale structure on plant-pollinator networks.

@@

# Data availability statement

All data and code is deposited at GitHub: https://github.com/JoseBSL/FunctionalMotifs.

>>>>

---
title: "The non-random assembly of indirect interactions in  \nplant-pollinator networks"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[left]{lineno}
- \linenumbers
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography_no_doi.bib
biblio-style: apalike
csl: JournalOfAnimalEcology.csl #converted from funct ecol style from zotero styles
---

__Jose B. Lanuza^1,2^__  | __Alfonso Allen-Perkins^2,3^__ | __Ignasi Bartomeus^2^__

1 School of Environmental and Rural Science, University of New England, Armidale, New South Wales 2350, Australia.

2 Estación Biológica de Doñana (EBD-CSIC), E-41092 Seville, Spain.

3 Departamento de Ingeniería Eléctrica, Electrónica, Automática y Física Aplicada, ETSIDI, Universidad Politécnica de Madrid, 28040 Madrid, Spain.

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organized is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. We found a higher proportion of densely connected motifs, which could reflect that neutral effects are the predominant ecological process governing the structure of network motifs. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different floral visitor and plant groups on both trophic levels. For example, bees tend to appear in motif positions with a lower number of indirect interactions than the rest of floral visitor groups. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlight the non-random structure of the key building blocks shaping plant-pollinators networks and the association of different floral visitor and plant groups with certain motifs.

Keywords: pollinators, plant-pollinator networks, plant-pollinator interactions, network motifs, indirect interactions, floral visitors

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions [i.e., the mediated effect between two species by a third species\; @holt1994], such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs [small sub-networks observed within a given network which are often referred as its building blocks\; @milo2002] allows to consider both direct and indirect interactions [@simmons2019]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches at the node or at the full network level are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network motifs can capture both direct and indirect interactions, by analysing over- and under-representation of motifs we were able to explore in great detail the structure of interactions between animal pollinators and the plants they visit. We have explored for the first time the overall structure and representation of network motifs in real plant-pollinator networks and shown that the different motifs up to five nodes are consistently over- or under-represented in comparison with simulated networks. Despite the presence of multiple ecological processes (e.g., neutral effects, morphological matching or phenological overlap) shaping motif structure, we found a tendency for over-represented structures with more densely connected motifs (i.e., complete, fan and asymmetric complete). As suggested in @simmons2020, a higher proportion of these densely connected motifs could reflect that neutral effects are the predominant ecological process governing the structure of network motifs. Thus, as indicated for other network properties like nestedness at the macro-scale level [@krishna2008; @suweis2013], or interaction probability at the species level [@bartomeus2016], species relative abundances (i.e., neutral effects) may be also the main driver governing the meso-scale structure on plant-pollinator networks.

@@

# Data Availability Statement

All data and code are publicly available at Github (https://github.com/JoseBSL/FunctionalMotifs) and Zenodo (https://doi.org/10.5281/zenodo.6489213).
